---
title: "weekly_july4"
author: "Dohun Lee"
date: '2022-07-20'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pracma); library(ggplot2); library(ggpubr); library(reshape2); 
library(mixtools); library(fitdistrplus); library(dplyr); library(QDNAseq); 
require(gtools); library(gridExtra); library(devtools)
source_url("https://raw.githubusercontent.com/elakatos/liquidCNA/main/mixture_estimation_functions.R")
knitr::opts_knit$set(root.dir = '/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/')
setwd("/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/")
source("SCRIPTS/3_liquidCNA.R")
load("../DATA/bam_by_patient.RData")
load("../DATA/antwerpData.RData")
load("../DATA/liquidCNA_results_july4.RData")
load("../DATA/new_ichor.RData")
```

This is a RMDmarkdown to track and show progress between 20/07/22 \~ 27/07/22.

There are four things I would like to share with you.

1. I am now having most of my absolute subclonality estimations in the range [0,1]. 
2. Boxplot suggested by Oscar. 
3. Timeplot of subclonality, purity and CT scan results mentioned by Solon.
4. liquidCNA vs ichorCNA (with results newly sent by Solon)

# 1. Absolute ratio [0,1]

```{r, include=FALSE}
RECIST <- append.RECIST()
```

```{r}
boxplot(RECIST$rat, 
        ylim = c(-0.5, 1.8),
        main = "Boxplot of all subclonality estimations",
        las = 1,
        col = c("lightblue"),
        border = "brown",
        horizontal = FALSE,
        notch = TRUE)
```

### I was able to achieve this by changing the baseline samples; instead of always setting it to Sample1, I set the Sample with the lowest mean CNS to be the baseline. 

# 2. Boxplot of subclonality for YES/NO Progression

### Subclonality: 

```{r}
#SUBCLONALITY X PROGRESSION!
na.or.base <- which(is.na(RECIST$rat) | RECIST$rat == 0)
RECIST.good <- RECIST[-na.or.base,]

yes.prog <- which(RECIST.good$Progression == "YES")

#
boxplot(RECIST.good[yes.prog,]$rat, 
        RECIST.good[-yes.prog,]$rat,
        ylim = c(0, 2),
        main = "Subclonality estimation against progression",
        at = c(1,2),
        names = c("YES", "NO"),
        las = 1,
        col = c("lightblue","pink"),
        border = "brown",
        horizontal = FALSE,
        notch = TRUE)
```

### Fortunately, the boxplot shows an increase in sub-clonality with progression. However, this is less apparent for purity.

### Purity: 

```{r}
#PURITY X PROGRESSION!
yes.prog <- which(RECIST.good$Progression == "YES")

#draw
boxplot(RECIST[yes.prog,]$purity_mean, 
        RECIST[-yes.prog,]$purity_mean,
        ylim = c(0, 0.6),
        main = "Purity estimation against progression",
        at = c(1,2),
        names = c("YES", "NO"),
        las = 1,
        col = c("lightblue","pink"),
        border = "brown",
        horizontal = FALSE,
        notch = TRUE)

```

# 3. Timeplot of subclonality, purity and CT scan result

### Here are two examples:

```{r}
plot.estimation.w.RECIST(2)
```

```{r}
plot.estimation.w.RECIST(7)
```

# 4. IchorCNA vs liquidCNA

There still aren't any apparent correlations between purity estimations of liquidCNA and ichorCNA. 

### **For ichorCNA, I am using the values under the column ichorCNA. Is this correct? What are the columns ACE, rascal & tMAD?**

```{r}
#get liquidCNA purities
purities <- unlist(sapply(1:length(liquidCNA_results), 
                          function(x) as.numeric(liquidCNA_results[[x]]$purity_mean)))

plot(purities, unlist(ichor.purities), xlab = "liquidCNA", ylab = "ichorCNA",
     main = "purity | liquidCNA vs ichorCNA")

```


