---
title: "weekly_june4"
author: "Dohun Lee"
date: '2022-06-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pracma); library(ggplot2); library(ggpubr); library(reshape2); 
library(mixtools); library(fitdistrplus); library(dplyr); library(QDNAseq); 
require(gtools); library(gridExtra); library(devtools)
source_url("https://raw.githubusercontent.com/elakatos/liquidCNA/main/mixture_estimation_functions.R")
knitr::opts_knit$set(root.dir = '/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/')
setwd("/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/")
source("SCRIPTS/3_liquidCNA.R")
load("../DATA/bam_by_patient.RData")
load("../DATA/antwerpData.RData")
load("../DATA/weekly_june4_data")
```

This is a RMDmarkdown to track and show progress between 22/06/22 \~ 30/06/22.

There were 3 main things to tackle:

1.  Cross checking BAM files with patients

-   TODO: identify erroneous bam file and send to Solon

2.  Retry batching of timely patients
3.  Retry 2 sample patients, trying comparison with base sample
4.  Understanding Equation 12

# 1. Cross referencing bam files

Checking if all BAM files are accounted for.
```{r BAM}
head(bam_by_patient)

#total number of bam data is equal between two
length(unname(unlist(bam_by_patient))) == nrow(ichorCNA)

#number of bam files assigned by me to each patient
vec_mine <- sapply(1:length(bam_by_patient), function(x) length(bam_by_patient[[x]]))

#number of ichorCNA data for each patient (TRUTH)
vec_ichor <- unname(table(sort(ichorCNA$Patient_ID)))

vec_mine==vec_ichor
```

# 2. Timely patients

There are 7 patients with 6+ BAM files:

Could do non overlapping batches, then stitch them together if absolute sub-clonal ratio can be extracted.

```{r six_plus}
sample_num_patient <- sapply(1:80, function(x) length(bam_by_patient[[x]]))
timely_patients <- which(sample_num_patient > 6)
timely_bam_num <- sample_num_patient[timely_patients]
```

### Problem 1: and error occurs in function normalmixEM() within estimateRGaussianFit()

Reason: when there's only one segment in seg.dcn.toUse, the function doesn't work. This is the case for Patient 1439.

Works well for some patients...

```{r}
res1[[5]]

```

NULL for patient 1439 with only one segment to use...

```{r}
res1[[2]]
```

**Problem/need addressing:** Absolute ratios are negative or over 1 for some patients. Additionally, some of the time samples will have a <NA> from not the EM not converging within 1000 steps. (Gaussian fitting step typically fails for low subclonal ratio sampes of <= 0.1)
```{r}
res1[[1]]

res1[[4]]
```


Questions:

* Why do some time samples have NA whilst others do not?
* What is an absolute ratio over 1, why does that happen? What is negative ratio?

Potential fix:

In the liquidCNA jupyter example, estimateRGaussFit function was called with: **estimateRGaussianFit(seg.dcn.toUse, samp, final.results, 3)**. However, I think this is a typo, and it is supposed to be **estimateRGaussianFit(seg.dcn.toUse, samp, final.results, nstates = 3)**. As the former sets the initial sigma for Gaussian as 3 for the EM algorithm. Fixing this code we get...:

# 3. 2 time patients

Changed the code in liquidCNA to output answer for patients with just two time samples.

* changed filterSegmentRatios to output a 1 column dataframe > vector for 2 time samples
* code for variable seg.dcn.toUse altered so that variable is a dataframe and not a vector

Relative subclonal ratio cannot be calculated as there is only 1 non-base time sample. However, absolute subclonal ratio can be calculated as a normal distribution is being fitted to dCN of segments. Only problem here this method can't yet choose an appropriate cutOff (i.e. deciding which segments are clonal or non-clonal, and furthermore, subclonal or unstable).

Trying the Gaussian fitting across all values of cutOff in cutOffVec for patient 1333. The results for one cutOff value (0.025) the final results table will be as following:

```{r }
eg_2samp_final.results
```

Absolute subclonal ratio across 96 cutOff values in cutOffVec. Why are some negative? Why do values repeat? How would we choose one?

```{r }
rats
table(rats)
```

# 4.  Understanding Equation 12

TODO: re read the paper












