---
title: "weekly_junly1"
author: "Dohun Lee"
date: '2022-07-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pracma); library(ggplot2); library(ggpubr); library(reshape2); 
library(mixtools); library(fitdistrplus); library(dplyr); library(QDNAseq); 
require(gtools); library(gridExtra); library(devtools)
source_url("https://raw.githubusercontent.com/elakatos/liquidCNA/main/mixture_estimation_functions.R")
knitr::opts_knit$set(root.dir = '/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/')
setwd("/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/")
source("SCRIPTS/3_liquidCNA.R")
load("../DATA/bam_by_patient.RData")
load("../DATA/antwerpData.RData")
load("../DATA/weekly_july1.RData")
load("../DATA/liquidCNA_results_july1.RData")
```

This is a RMDmarkdown to track and show progress between 30/06/22 \~ 06/07/22.

There were 3 main things to tackle:

1.  Reproduce figure 3C
2.  Make estimateRGaussian() work with nstates = 3
  +   Check why there is an error for patient 1439
  +   Is there really a problem if EM requires more than 800 steps? Try increasing this. -> 1500, got rid of all NAs, what are its implications
  +   reRun for all patients can collect results for all -> DONE
    - Record nrow(seg.dcn.toUse)
3.  Append purity and subclonal ratios of each patient-time sample onto RECIST


# 1. Figure 3C

Figure 3C is showing...: *"Copy number profile of sample, with raw bin-wise and segmented copy number values shown in black and red, respectively."*

Can we colour in the segments that liquidCNA designates to be sub-clonal and is used to estimate sub-clonal ratio?

**Problem 1: the segments used in liquidCNA are ensemble segments, so are subsets of segments identified in QDNAseq.**

Answer 1: we do have start and end (bin number, i.e., row of seg.df) for ensemble segments
```{r}
head(patient1.seg.data)
```

**Problem 2: How can I parse through a QDNAseq data object? Can data regarding ensemble segments be appended then plotted?**

```{r QDNAseqObject}
copyNumbersSegmented
plot(copyNumbersSegmented)
```


# 2. Run liquidCNA for all patients and study the results

## Run for normal patients

```{r}
#Patients with only two samples
sample_num_patient <- sapply(1:80, function(x) length(bam_by_patient[[x]]))
two_samp_patients <- which(sample_num_patient == 2)

#Patients with over 6 samples
timely_patients <- which(sample_num_patient > 6)

go_patients <- 1:80
go_patients <- go_patients[!(go_patients %in% two_samp_patients)]
go_patients <- go_patients[!(go_patients %in% timely_patients)]
```

Run!
```{r, eval = FALSE, include=TRUE}
#initialise results vector
go_results <- vector(mode = "list", length = length(go_patients))

#Run for go_patients (i.e., non 2 or timely sample patients)
for(patient_x in 1:length(go_patients)){
  tryCatch({
    cat("Starting patient :", patient_x, "\n")
    go_results[[patient_x]] <- run_liquidCNA(go_patients[patient_x])
  }, error=function(e){cat("ERROR at:", go_patients[patient_x], "\n")})
}

names(go_results) <- paste0("patient_",patient_ids[go_patients])
```

### GO: Investigate error
Investigate: 

*   patients with NULL results,
*   patients with NA for absolute value,
*   patients with absolute ratio out of boundary of 0~1. <- in fact, rat is absolute ratio so a rat value outside of boundary would be fine? as negative value means less subclones than base sample, and rat > 1 is simply that much more than the base sample?

```{r}
#NULL results
null_go <- go_patients[which(sapply(1:length(go_patients), function(x) is.null(go_results[[x]])))]
null_go

#NA absolute - all time samples have an absolute ratio
any(sapply(1:length(go_patients), function(x) any(is.na(go_results[[x]]$rat))))

#absolute smaller than 0 - NONE
any(sapply(1:length(go_patients), function(x) any(as.numeric(go_results[[x]]$rat) < 0)))

#absolute greater than 1 - absolute ratio so bigger than 1 is fine?
big_rat_go <- which(sapply(1:length(go_patients), function(x) any(as.numeric(go_results[[x]]$rat) > 1)))
  #ids of patients with rat > 1
patient_ids[go_patients[big_rat_go]]
```

Tackle NULL patients, where is the ERROR coming from? and how can we fix it?

## Run for two sample patients

Run!
```{r,eval = FALSE, include=TRUE}
#initialise results vector
two_samp_results <- vector(mode = "list", length = length(two_samp_patients))

#run!
for(patient_x in 1:length(two_samp_patients)){
  tryCatch({
    cat("Starting patient :", patient_x, "\n")
    two_samp_results[[patient_x]] <- run_2samp_liquidCNA(two_samp_patients[patient_x])
  }, error=function(e){cat("ERROR at:", patient_x, "\n")})
}

names(two_samp_results) <- paste0("patient_",patient_ids[two_samp_patients])
```


```{r}
head(two_samp_results)
```

## Run for 6+ patients

```{r, eval=FALSE, include=TRUE}
#number of BAM files our timely patients have
timely_bam_num <- sample_num_patient[timely_patients]

#number of bam files to batch each timely patient by:
nbatch1 <- ceiling(timely_bam_num/2)
nbatch2 <- timely_bam_num - nbatch1

#initialise result vector
timely_res1 <- vector(mode = "list", length = length(timely_patients))
timely_res2 <- vector(mode = "list", length = length(timely_patients))

#RUN!
for(x in 1:length(timely_patients)){
  tryCatch({
    cat("\n Starting", x, "\n")
    timely_res1[[x]] <- run_liquidCNA(timely_patients[x], 
                               timely = TRUE,
                               batch1 = TRUE,
                               nbatch = nbatch1[x])
  }, error=function(e){cat("ERROR at:", x, "\n")})
}

for(x in 1:length(timely_patients)){
  tryCatch({
    cat("\n Starting", x, "\n")
    timely_res2[[x]] <- run_liquidCNA(timely_patients[x], 
                               timely = TRUE,
                               batch1 = FALSE,
                               nbatch = nbatch2[x])
  }, error=function(e){cat("ERROR at:", x, "\n")})
}

names(timely_res1) <- paste0("patient_",patient_ids[timely_patients],"_batch1")
names(timely_res2) <- paste0("patient_",patient_ids[timely_patients],"_batch2")

```


### Timely: Investigate error

I think this may be due to erroneous BAM file present for this patient - not sure why though...

Erroneous BAM file: SLX.17921.D701tp_D505tp.H5J55BBXY.s_8.r_1
```{r}
error_seg.df <- read.delim(paste0("../DATA/2_QDNA_CNout/segment_2789_og.txt"))
error_seg.df <- error_seg.df[,-(1:4)]
colnames(error_seg.df) <- paste0("Sample", 1:ncol(error_seg.df))
head(error_seg.df)
```

### Stitch Timely

Question - for the results, how are the rows ordered? should i reorder them so they just go up? this won't take long just look
A - in final.results, rows are ordered in the sample order found by liquidCNA with base sample appended in the final row

**Problem: this method of stitching doesn't make sense. as the assumption of liquidCNA, results the base sample of second batch to have sub-clonality of 0**
```{r}
batch1_rat <- as.numeric(timely_res1[[1]]$rat)
batch2_rat <- as.numeric(timely_res2[[1]]$rat)

batch1_rat
batch2_rat

rat_ordered <- c(batch1_rat[length(batch1_rat)], batch1_rat[1:(length(batch1_rat)-1)], batch2_rat[length(batch2_rat)], batch2_rat[1:(length(batch2_rat)-1)])

plot(rat_ordered, type = "l", ylab = "Absolute sub-clone ratio estimated",
     xlab = "Ordered Samples")
```























