---
title: "weekly_july5"
author: "Dohun Lee"
date: '2022-07-29'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pracma); library(ggplot2); library(ggpubr); library(reshape2); 
library(mixtools); library(fitdistrplus); library(dplyr); library(QDNAseq); 
require(gtools); library(gridExtra); library(devtools)
library("dplyr")
library("ggpubr")
source_url("https://raw.githubusercontent.com/elakatos/liquidCNA/main/mixture_estimation_functions.R")
knitr::opts_knit$set(root.dir = '/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/')
setwd("/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/")
source("SCRIPTS/3_liquidCNA.R")
load("../DATA/bam_by_patient.RData")
load("../DATA/antwerpData.RData")
load("../DATA/liquidCNA_results_july4.RData")
```

Hello Solon, this is a Rmarkdown to discuss two of the things I've been working on:

1.  Two group statistical testings
2.  KM curves

# 1.0 P-values

First, append purity and subclonality estimations from liquidCNA. **Then, group subclonalities depending on Progression is YES or NO.**
```{r}
RECIST <- append.RECIST()
yes.prog <- which(RECIST$Progression == "YES")
no.prog <- which(RECIST$Progression == "NO")

yes.progression <- RECIST[yes.prog,]$rat
no.progression <- RECIST[no.prog,]$rat

no.progression <- no.progression[!(is.na(no.progression))]
yes.progression <- yes.progression[!(is.na(yes.progression))]
```

### T-test:

First, using shapiro to test for normality:
```{r}
shapiro.test(no.progression) # p-value < 0.05
shapiro.test(yes.progression) # p-value < 0.05
```

p-value is less than 0.05 implying that the distribution of the data are significantly different from normal distribution. Thus, the distribution of estimated subclonalities are skewed, **normality should not be assumed and T-test is inappropriate**.

Still, doing the t-test, p-value is 0.3489
```{r}
t.test(no.progression, yes.progression, var.equal = FALSE) #p-value > 0.05
```

As normality cannot be assumed, Unpaired Two-Samples **Wilcoxon Test** was conducted instead. **This however resulted a p-value of 1**
```{r}
wilcox.test(no.progression, yes.progression, alternative = "two.sided")
```

As opposed to this **purity estimation results significant for both** T-test and Wilcox test:

```{r}
t.test(RECIST[no.prog,]$purity_mean, RECIST[yes.prog,]$purity_mean, var.equal = FALSE)

wilcox.test(RECIST[no.prog,]$purity_mean, RECIST[yes.prog,]$purity_mean, alternative = "two.sided")

```

**With Wilcox testing giving a p-value of 1, I am wondering if I have made an error in doing the statistic test somewhere. Do you have any idea what may be the cause here?**

# 2.0 Using subclonality to estimate metastasis

For each patient, they were grouped into whether they have any new metastasis ('y' in RECIST) or not. Then, for each group, their respective subclonality estimations were grouped as well:

```{r}
#function to categorise patient as newly metastasised or not

patient.metastasis <- function(p.id){
  p.RECIST <- RECIST[which(RECIST$Patient_ID == p.id),]
  if(p.RECIST$time[1] <= 10){
    p.RECIST <- p.RECIST[-1,]
  }
  m <- "y" %in% unname(unlist(p.RECIST[,5:14]))
  return(m)
}

#use the function and group...
p.metastasis <- sapply(unique(RECIST$Patient_ID), function(x) patient.metastasis(x))

#group subclonality estimations respectively....
subclonalities <- (sapply(1:length(liquidCNA_results), 
                    function(x) as.numeric(liquidCNA_results[[x]]$rat)))
yes.met.rat <- unlist(sapply(which(p.metastasis), function(x) subclonalities[[x]]))
no.met.rat <- unlist(sapply(which(p.metastasis==F), function(x) subclonalities[[x]]))

ts.each.metastasis
ts.yes.met.rat <-RECIST[ts.each.metastasis,]$rat
ts.no.met.rat <-RECIST[ts.each.metastasis==F,]$rat
```

Doing the Wilcox test, **we get a significant result**. I.e., distribution of subclonalities between patient groups of further metastasis and not, significantly differs:

```{r}
p.met.res <- wilcox.test(yes.met.rat, no.met.rat, alternative = "two.sided")
ts.met.res <- wilcox.test(ts.yes.met.rat, ts.no.met.rat, alternative = "two.sided")
boxplot(no.met.rat, 
        yes.met.rat,
        ylim = c(0, 3),
        main = "Subclonality against new metastasis (by patient)",
        at = c(1,2),
        names = c("NO", "YES"),
        las = 1,
        col = c("pink","lightblue"),
        border = "brown",
        horizontal = FALSE,
        notch = TRUE,
        ylab = "Subclonal ratio estimation")
text(2.1, 2.9, paste0("Wilcox test: \np-value = ",round(p.met.res$p.value, 5)))
```

Thus, survival analysis was done next

# 3.0 Survival analysis

load Data and package
```{r}
library("survival")
library("survminer") 
load("../DATA/survival_df.RData")
```

Survival.df
```{r}
head(survival.df2)
```

1.  The first column are patient ids
2.  tstart, tstop are in days of event. **Days start from RECIST's "Date.Meta" column. Could you confirm if this is the date of first primary metastasis for the patient please?**
3.  status: 1 is recurrent event; 0 is right censored data
4.  Event: 2 is metastasis; 1 is no event

Furthermore, there are three covariates included in the data frame: sub.group (whether level of subclonality is high (2) or low (1)), subclonality ratio estimate and purity estimate. For sub.group,  2 is high subclonality group; and 1 is low subclonality group. Group2 has subclonality > 0.3656956 and 1 below it.

### KM was done first:

```{r}
model.2 = survfit(Surv(tstart,tstop,status) ~ sub.group,  data = survival.df2)
ggsurvplot(model.2, data = survival.df2,
     xlab = "Days", 
     ylab = "Overall survival probability")
```

**The problem with KM is that events are single-off. So even if metastasis (our event in question) reoccurs, the analysis doesnot take this into account when computing. Therefore, Cox proportional hazards regression model (which allows for recurrent events) was explored. **

The three covariates were used to model the Cox regression. Again, **purity was the only factor coming out to be significant**

```{r}
model.1 = coxph(Surv(tstart,tstop,status) ~ (sub.group + subclone + purity + cluster(id)), 
                method="breslow", 
                robust=TRUE, 
                data = survival.df2)
summary(model.1)

```

**Plotting the survival(metastasis) probability...**
```{r}
ggsurvplot(survfit(model.1), data = survival.df2)
```

**Plotting the cumulative hazard rate...**
```{r}
sfit <- survfit(model.1)
cumhaz.upper <- -log(sfit$upper)
cumhaz.lower <- -log(sfit$lower)
cumhaz <- sfit$cumhaz # same as -log(sfit$surv)

plot(cumhaz, xlab="Days ahead", ylab="cumulative hazard",
     ylim=c(min(cumhaz.lower), max(cumhaz.upper)))
lines(cumhaz.lower)
lines(cumhaz.upper)

```


# From Solon's feedback: a Test based on binomial distribution

Your null hypothesis is: the average subclonality proportion for Progression = Yes is equal to the average subclonality proportion for Progression = No. 

```{r}
yes.prog <- which(RECIST$Progression == "YES")
no.prog <- which(RECIST$Progression == "NO")

yes.progression <- RECIST[yes.prog,]$rat
no.progression <- RECIST[no.prog,]$rat

no.progression <- no.progression[!(is.na(no.progression))]
yes.progression <- yes.progression[!(is.na(yes.progression))]

prop.test(x = c(mean(no.progression)*100, mean(yes.progression)*100),
          n = c(length(no.progression), length(yes.progression)))

binom.test(x, n, p)
```

















