---
title: "weekly_july3"
author: "Dohun Lee"
date: '2022-07-14'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pracma); library(ggplot2); library(ggpubr); library(reshape2); 
library(mixtools); library(fitdistrplus); library(dplyr); library(QDNAseq); 
require(gtools); library(gridExtra); library(devtools)
source_url("https://raw.githubusercontent.com/elakatos/liquidCNA/main/mixture_estimation_functions.R")
knitr::opts_knit$set(root.dir = '/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/')
setwd("/Users/dhl/laheen Dropbox/DHL DHL/Cambridge/Internship/subclonality/")
source("SCRIPTS/3_liquidCNA.R")
load("../DATA/bam_by_patient.RData")
load("../DATA/antwerpData.RData")
load("../DATA/weekly_july3.RData")
load("../DATA/liquidCNA_results_july1.RData")
load("../DATA/liquidCNA_results_july2.RData")
```

This is a RMDmarkdown to track and show progress between 13/07/22 \~ 27/07/22.

These were the main things to tackle:

1.  Absolute ratio 0~1
2.  ichorCNA: correlation between liquidCNA and ichorCNA's optimichorCNA & CTC_count
3.  Compare the absolute ratios between timely batch 2 options
4.  RECIST: look at metastatic data; look at progression; evaluate whether subclonality tells us anything; compare subclonality between Y and N metastasis

# 1. Absolute ratio 0~1

### Patient 3080 | patient with extreme absolute sub-clonality estimation:

```{r}

liquidCNA_results$patient_3080

```

### I think the extreme absolute subclonality ratio is coming from extreme deltaCN values

```{r}
head(seg.dcn.toUse.46)
```
### The extreme deltaCN seems to come from low purity...

### The three time samples have the following purity estimates:
```{r}
pVec.46
```

### This results the sample with low purity to have extreme CNS when corrected by purity.

### Before correction the segment CN values look normal...
```{r}
head(seg.cns.46)
```
### However, with correction...
```{r}
head(seg.cns.corr.46)
```
### The equation they correct CNS with purity by: **(((CNS-2)*1/purity)+2)**

### If we correct the CNS value of the first segment for each sample... 
```{r}
(1.371152	 - 2)/0.05 + 2 #Sample1
(2.004080 - 2)/0.07 + 2 #Sample2
(1.730489 - 2)/0.13 + 2 #Sample3
```

### This results dCN to be extreme

```{r}
head(seg.dcn.46)
```

### So it seems that the extreme subclonality estimates are coming from low purity and purity-corrected CNS rather than the GMM fitting step

## Take the samples where sub-ratio is outside [0,1] and check the corresponding purity: 

Majority of the error looks to come from small purity values. Currently, 31 out of 283 estimates are outside [0,1]; this is from 22/80 patients. Additionally, only 23 of the estimates linked with RECIST are outside the boundary.

```{r}
rats <- unlist(sapply(1:length(liquidCNA_results), 
                      function(x) as.numeric(liquidCNA_results[[x]]$rat)))
purities <- unlist(sapply(1:length(liquidCNA_results), 
                          function(x) as.numeric(liquidCNA_results[[x]]$purity_mean)))

# 31/283 estimations have sub-clonality ratio outside of [0,1]
out.rats <- which(rats < 0 | rats > 1)
length(out.rats)

is.out.rat <- function(x){
  rats <- as.numeric(liquidCNA_results[[x]]$rat)
  return(any(rats < 0 | rats > 1))
}

# 22/80 patients have sub-clonality ratio outside of [0,1]
out.rat.patients <- which(sapply(1:length(liquidCNA_results), function(x) is.out.rat(x)))
length((out.rat.patients))

#purities of out.rats
rats[out.rats]
purities[out.rats]

plot(purities[out.rats], rats[out.rats],
     ylab = "Sub-ratio estimate outside [0,1]",
     xlab = "Corresponding purity estimates")

out.results <- sapply(out.rat.patients, function(x) liquidCNA_results[[x]], simplify = F)
names(out.results) <- paste0("patient_", patient_ids[out.rat.patients])
```

Feedback: "it would be nice to see what causes the high-purity samples in your plot to fail as well? You can focus on the 2-3 with the highest purity and investigate them further."

## Investigating why purities are failing:

### idea 1: when one of the samples have drastically lower purity
```{r}
out.results$patient_3614
```

Purity values:
```{r}
pVec.58
```

Again, the dCN values for the erroneous sample are at a greater scale.
```{r}
head(seg.dcn.toUse.58)
```

### If we look at segment CN values, we can see that the CN states are at a drastically different range. This is resulted during the renomarlisation stage (where the centre of CN is recentred to be exactly 2). 

Could we trim them out? not consider these segments? I tried setting the maximum CN value to be 6 (as the CN states considered during purity estimation are 5 CN states anyways). Whilst this method did not 'save' the result for patient 3614, it did bring down the absolute subclonality estimation to be near 1 (so I believe it would save the result for less drastic patients). One thing is, this seems like I am disturbing high quality sample (high purity) to fit well with lower quality samples (low purity).

```{r}
plot.3C(3, cn.df.58, seg.df.58, seg.sub.58, seg.plot.58, 3301)
plot.3C(2, cn.df.58, seg.df.58, seg.sub.58, seg.plot.58, 3301)
```

```{r}
plot(density(seg.df.58[,2]), main = "density plot of Sample 2's seg.df")
plot(density(seg.df.58[,3]), main = "density plot of Sample 3's seg.df")

plot(density(cn.df.58[,2]), main = "density plot of Sample 2's cn.df")
plot(density(cn.df.58[,3]), main = "density plot of Sample 3's cn.df")
```


# 2. ichorCNA

### ichorCNA prediction of purity versus liquidCNA

There doesn't seem to be much correlation between the two estimations. Of note, for many of the samples, one of the algorithms estimate purity = 0, whilst the other predict a wide range of purities. This is expected as (in liquidCNA) baseline samples are restricted/assumed to purity of 0. 

```{r}
optimichor <- ichorCNA$OptimichorCNA
ctc_count <- ichorCNA$CTC_count

ichorCNA.sorted <- ichorCNA[order( ichorCNA$Patient_ID, ichorCNA$Date ),]

ichor.purities <- ichorCNA.sorted$OptimichorCNA
ichor.purities <- ichor.purities / 100

plot(purities, ichor.purities, xlab = "liquidCNA", ylab = "ichorCNA",
     main = "purity | liquidCNA vs ichorCNA")
```

### Circulating tumor cell (CTC) count

```{r}
plot(ichor.purities, ctc_count, xlab = "ichorCNA purity", ylab = "ichorCNA CTC",
     main = "liquidCNA purity vs ichorCNA CTC count")
```


# 4. Linking with RECIST 

```{r, include=FALSE}
build.df <- function(patient_id){
  #get patient's liquidCNA results, RECIST and ichorCNA data
  p_num <- match(patient_id, patient_ids)
  p.liquid <- liquidCNA_results[[p_num]]
  p.RECIST <- RECIST[which(RECIST$Patient_ID == patient_id),]
  p.ichor <- ichorCNA[which(ichorCNA$Patient_ID == patient_id),]
  
  p.rat.last <- tail(p.RECIST$rat, n=1) 
  p.rat.second <- p.RECIST$rat[2]
  p.purity.last <- tail(p.RECIST$purity_mean, n=1) 
  p.purity.first <- p.RECIST$purity_mean[1]
  
  if(nrow(p.RECIST)==0){
    p.df <- data.frame(p.progression = NA,
                       OS = NA,
                       purity.increasing = NA,
                       purity.change = NA,
                       subclone.increasing = NA,
                       subclone.change = NA)
    rownames(p.df) <- paste0("patient_", patient_id)
    return(p.df)
  } else {
    
    #1. whether there is a change in progression
      # 1 is always progression
      # 0 is always stable
      # -1 is mix of progression and stable
    progression <- ifelse(p.RECIST$Progression=="YES",1,0)
    any.na <- any(is.na(progression))
    test <- sum(progression) - length(progression)
    always.stable <- ifelse(test == -(length(progression)), 1, 0)
    if(any.na){
      p.progression <- NA
    } else {
      if(sign(test) == 0){
        p.progression <- 1
      } else {
        p.progression <- ifelse(always.stable == 1, 0, -1)
      }
    }
    
    #2. patient alive? 1 is yes, 0 is deceased
    OS <- ifelse(all(p.ichor$Deceased == "alive"), 1, 0)
    
    #3. Purity Boolean. purity, is overall increasing or decreasing over time samples?
    #boolean of is increasing? 1 is yes, 0 is no (decreasing)
    purity.increasing <- ifelse((p.purity.last - p.purity.first)>0, 1, 0)
    
    #4. Purity. how much does purity change between first blood draw and last?
    purity.change <- p.purity.last - p.purity.first
    
    #5. Sub-clone Boolean. whether sub-clonality overall decreased or increased over time samples.
    #boolean of is increasing? 1 is yes, 0 is no (decreasing)
    subclone.increasing <- ifelse((p.rat.last - p.rat.second)>0, 1, 0)
    
    #6. Subclone. absolute change in sub-clonality between first blood draw and last:
    subclone.change <- p.rat.last
    
    #package into a dataframe
    p.df <- data.frame(p.progression = p.progression,
                       OS = OS,
                       purity.increasing = purity.increasing,
                       purity.change = purity.change,
                       subclone.increasing = subclone.increasing,
                       subclone.change = subclone.change)
    rownames(p.df) <- paste0("patient_", patient_id)
    return(p.df)
    
  }
}
```

For each patient, build.df() builds a dataframe with 6 information:

  1. Whether there is a change in progression
    + 1 is always progression
    + 0 is always stable
    + -1 is mix of progression and stable
  2. Patient alive? 1 is yes, 0 is deceased
  3. Purity Boolean. purity, is overall increasing or decreasing over time samples? Boolean of is increasing? 1 is yes, 0 is no (decreasing)
  4. Purity. how much does purity change between first blood draw and last?
  5. Sub-clone Boolean. whether sub-clonality overall decreased or increased over time samples. boolean of is increasing? 1 is yes, 0 is no (decreasing)
  6. Subclone. absolute change in sub-clonality between first blood draw and last.
  
```{r}
head(c.df)
```












